---
title: "Codebook: Utility Efficiency, Market Structure, and Access to Electricity"
author: "Edgar Aguilar"
date: "April 20, 2025"
output:
  pdf_document:
    toc: true
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
df <- read_csv("Efficiency_Markets_Access.csv")
```
\newpage
# Overview of the Data

This codebook accompanies the dataset "Efficiency_Markets_Access.csv", which merges utility-level performance metrics with country-level electricity market structure, governance characteristics, and national electricity access rates. The dataset supports a benchmarking analysis of 49 electric utilities across 37 countries and includes 21 cleaned and standardized variables.

The data was processed through multiple steps, including cleaning, merging, reshaping, and the construction of efficiency scores using Data Envelopment Analysis (DEA). These steps are fully documented in the replication script "Efficiency_scores-market-access_dataset_creation.R", which ensures the process is transparent and reproducible.

# Sources and Data Wrangling

**Data Sources:**

- [UPBEAT Utility Database](https://utilityperformance.energydata.info/about/intro): Utility performance and financial indicators.
- [World Bank Market Structure Dataset](https://datacatalog.worldbank.org/search/dataset/0065245): Country-level electricity market structures and governance characteristics.
- [World Bank - Access to Electricity (WDI)](https://data.worldbank.org/indicator/EG.ELC.ACCS.ZS): National electricity access rates.

**Wrangling Steps:**

- Computed the average value of input and output indicators across all available years for each utility.
- Performed DEA using an input-oriented CCR model in R.
- Merged DEA scores with 2012 market structure data.
- Joined 2022 national electricity access data.
- Standardized names and cleaned structure.

# Data Envelopment Analysis (DEA) Methodology

The **Efficiency Score** variable in this dataset was constructed using **Data Envelopment Analysis (DEA)**, a non-parametric technique for benchmarking the relative efficiency of utilities in transforming inputs into outputs. This project applies an **input-oriented CCR model** (assuming constant returns to scale), which estimates how much a utility could reduce its inputs while maintaining its output levels.

The input and output variables were sourced from the **UPBEAT utility performance database**, a World Bank initiative that standardizes multi-year utility data across financial, operational, and technical indicators. For each utility, input and output values were averaged across all available years in the dataset to create a representative profile.

**Inputs:**
- Employees
- Total Assets

**Outputs:**
- Net Profit Margin
- Customers
- Transformed Distribution Losses (1 - Distribution Losses)

The model was implemented in R using the `Benchmarking` package. Utilities with missing input/output values were excluded.

# Variable Descriptions and Summaries

## Utility Long Name
- **Type:** Character
- **Description:** Full legal name of the electric utility.
- **Source:** UPBEAT

## Utility Short Name
- **Type:** Character
- **Description:** Abbreviated or acronym version of the utility name.
- **Source:** UPBEAT

## Ownership
- **Type:** Factor
- **Description:** Ownership structure of the utility (Public or Private).
- **Source:** UPBEAT
```{r}
table(df$Ownership, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Ownership", "Count"),
    caption = "Ownership Structure of Utilities"
  ) %>%
  kable_styling(full_width = FALSE)
```

## Type
- **Type:** Factor
- **Description:** Categorical variable describing the operational structure of the utility. All utilities in the dataset include a **distribution component** in their operations, but may also engage in other segments of the power sector:
  - **D (Distribution-only):** Utility is solely responsible for electricity distribution.
  - **G&D (Generation and Distribution):** Utility is involved in both electricity generation and distribution, but not transmission.
  - **VIU (Vertically Integrated Utility):** Utility is active across the entire electricity value chain, including generation, transmission, and distribution.
- **Source:** UPBEAT
```{r}
table(df$Type, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Type", "Count"),
    caption = "Distribution of Utility Operational Structures"
  ) %>%
  kable_styling(full_width = FALSE)
```

\newpage
## Country
- **Type:** Character
- **Description:** Country of operation.
- **Source:** World Bank Market Structure Dataset
```{r country-list-kable-wrap}
# Create comma-separated, wrapped country list
country_list <- df %>%
  distinct(Country) %>%
  arrange(Country) %>%
  pull(Country) %>%
  paste(collapse = ", ")

# Display table
tibble(Countries = country_list) %>%
  kable(caption = "List of Countries in the Dataset", booktabs = TRUE, escape = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = "hold_position") %>%
  column_spec(1, width = "13cm")
```
*The dataset includes `r n_distinct(df$Country)` unique countries.*

## Region  
- **Type:** Factor
- **Description:** Geographical region classification assigned to each country. 
- **Source:** UPBEAT (World Bank classification)
```{r}
table(df$Region, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Region", "Count"),
    caption = "Distribution of Utilities by Geographical Region"
  ) %>%
  kable_styling(full_width = FALSE)
```

## Income Group
- **Type:** Factor
- **Description:** Country income classification (Low, Lower-middle, etc.).
- **Source:** UPBEAT (World Bank classification)
```{r}
table(df$`Income Group`, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Income Group", "Count"),
    caption = "Distribution of Utilities by Income Group"
  ) %>%
  kable_styling(full_width = FALSE)
```

\newpage
## Employees
- **Type:** Numeric
- **Description:** Total number of employees in utility company. 
- **Source:** UPBEAT
```{r}
df %>%
  summarize(
    Min = min(Employees, na.rm = TRUE),
    Q1 = quantile(Employees, 0.25, na.rm = TRUE),
    Median = median(Employees, na.rm = TRUE),
    Mean = mean(Employees, na.rm = TRUE),
    Q3 = quantile(Employees, 0.75, na.rm = TRUE),
    Max = max(Employees, na.rm = TRUE),
    NAs = sum(is.na(Employees))
  ) %>%
  kable(caption = "Summary - Employees", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Total Assets
- **Type:** Numeric
- **Description:** Total assets in USD millions.
- **Source:** UPBEAT
```{r}
df %>%
  summarize(
    Min = min(`Total Assets`, na.rm = TRUE),
    Q1 = quantile(`Total Assets`, 0.25, na.rm = TRUE),
    Median = median(`Total Assets`, na.rm = TRUE),
    Mean = mean(`Total Assets`, na.rm = TRUE),
    Q3 = quantile(`Total Assets`, 0.75, na.rm = TRUE),
    Max = max(`Total Assets`, na.rm = TRUE),
    NAs = sum(is.na(`Total Assets`))
  ) %>%
  kable(caption = "Summary - Total Assets", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Net Profit Margin
- **Type:** Numeric
- **Description:** Net profit as a percentage of revenue.
- **Source:** UPBEAT
```{r}
df %>%
  summarize(
    Min = min(`Net Profit Margin`, na.rm = TRUE),
    Q1 = quantile(`Net Profit Margin`, 0.25, na.rm = TRUE),
    Median = median(`Net Profit Margin`, na.rm = TRUE),
    Mean = mean(`Net Profit Margin`, na.rm = TRUE),
    Q3 = quantile(`Net Profit Margin`, 0.75, na.rm = TRUE),
    Max = max(`Net Profit Margin`, na.rm = TRUE),
    NAs = sum(is.na(`Net Profit Margin`))
  ) %>%
  kable(caption = "Summary - Net Profit Margin", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Customers
- **Type:** Numeric
- **Description:** Number of customers served by utility company.
- **Source:** UPBEAT
```{r}
df %>%
  summarize(
    Min = min(Customers, na.rm = TRUE),
    Q1 = quantile(Customers, 0.25, na.rm = TRUE),
    Median = median(Customers, na.rm = TRUE),
    Mean = mean(Customers, na.rm = TRUE),
    Q3 = quantile(Customers, 0.75, na.rm = TRUE),
    Max = max(Customers, na.rm = TRUE),
    NAs = sum(is.na(Customers))
  ) %>%
  kable(caption = "Summary - Customers", align = "c") %>%
  kable_styling(full_width = FALSE)
```

\newpage
## Distribution Losses
- **Type:** Numeric
- **Description:** Share of electricity lost in distribution.
- **Source:** UPBEAT
```{r}
df %>%
  summarize(
    Min = min(`Distribution Losses`, na.rm = TRUE),
    Q1 = quantile(`Distribution Losses`, 0.25, na.rm = TRUE),
    Median = median(`Distribution Losses`, na.rm = TRUE),
    Mean = mean(`Distribution Losses`, na.rm = TRUE),
    Q3 = quantile(`Distribution Losses`, 0.75, na.rm = TRUE),
    Max = max(`Distribution Losses`, na.rm = TRUE),
    NAs = sum(is.na(`Distribution Losses`))
  ) %>%
  kable(caption = "Summary - Distribution Losses", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Transformed Distribution Losses
- **Type:** Numeric
- **Description:** 1 minus distribution losses.
- **Source:** UPBEAT (calculated)
- **Note:** This is a transformation of "Distribution Losses" variable to fit DEA model appropiately as an output.
```{r}
df %>%
  summarize(
    Min = min(`Transformed Distribution Losses`, na.rm = TRUE),
    Q1 = quantile(`Transformed Distribution Losses`, 0.25, na.rm = TRUE),
    Median = median(`Transformed Distribution Losses`, na.rm = TRUE),
    Mean = mean(`Transformed Distribution Losses`, na.rm = TRUE),
    Q3 = quantile(`Transformed Distribution Losses`, 0.75, na.rm = TRUE),
    Max = max(`Transformed Distribution Losses`, na.rm = TRUE),
    NAs = sum(is.na(`Transformed Distribution Losses`))
  ) %>%
  kable(caption = "Summary - Transformed Distribution Losses", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Operating and Debt Service Cost Recovery
- **Type:** Numeric
- **Description:** Ratio of billed revenue to cost of service, calculated on a billed basis (excluding subsidies and collections). This reflects whether the utility is generating enough billed revenue to cover its operating costs and debt service obligations.
- **Source:** UPBEAT
```{r}
df %>%
  summarize(
    Min = min(`Operating and Debt Service Cost Recovery`, na.rm = TRUE),
    Q1 = quantile(`Operating and Debt Service Cost Recovery`, 0.25, na.rm = TRUE),
    Median = median(`Operating and Debt Service Cost Recovery`, na.rm = TRUE),
    Mean = mean(`Operating and Debt Service Cost Recovery`, na.rm = TRUE),
    Q3 = quantile(`Operating and Debt Service Cost Recovery`, 0.75, na.rm = TRUE),
    Max = max(`Operating and Debt Service Cost Recovery`, na.rm = TRUE),
    NAs = sum(is.na(`Operating and Debt Service Cost Recovery`))
  ) %>%
  kable(caption = "Summary - Cost Recovery", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Efficiency Score
- **Type:** Numeric
- **Description:** A Data Envelopment Analysis (DEA) efficiency score ranging from 0 to 1, where 1 indicates a fully efficient utility. Scores were computed using an input-oriented CCR model based on selected input (Employees, Total Assets) and output (Net Profit Margin, Customers, Transformed Distribution Losses) variables.
- **Source:** UPBEAT (DEA computed)
```{r}
df %>%
  summarize(
    Min = min(`Efficiency Score`, na.rm = TRUE),
    Q1 = quantile(`Efficiency Score`, 0.25, na.rm = TRUE),
    Median = median(`Efficiency Score`, na.rm = TRUE),
    Mean = mean(`Efficiency Score`, na.rm = TRUE),
    Q3 = quantile(`Efficiency Score`, 0.75, na.rm = TRUE),
    Max = max(`Efficiency Score`, na.rm = TRUE),
    NAs = sum(is.na(`Efficiency Score`))
  ) %>%
  kable(caption = "Summary - Efficiency Score", align = "c") %>%
  kable_styling(full_width = FALSE)
```

## Market Structure  
- **Type:** Factor  
- **Description:** Categorical variable representing the utility's national power market structure, based on a simplified classification derived from World Bank market reform data. The three categories are:
  - **VIU (Vertically Integrated Utility):** The utility operates within a fully integrated system controlling generation, transmission, and distribution.
  - **SBM (Single Buyer Model):** A central purchasing entity procures electricity from generators and sells it to distributors or end-users.
  - **WRC (Wholesale-Retail Competition):** Market features competition at both the wholesale and retail levels, allowing multiple sellers and/or buyers.
- **Note:** This classification reflects the extent of market liberalization and unbundling reforms in the utility’s country of operation as of 2012.
- **Source:** World Bank Market Structure Dataset
```{r}
table(df$`Market Structure`, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Market Structure", "Count"),
    caption = "Distribution of Utilities by Market Structure"
  ) %>%
  kable_styling(full_width = FALSE)
```

## Regulator
- **Type:** Logical
- **Description:** TRUE if an independent regulator exists.
- **Source:** World Bank Market Structure Dataset
```{r}
table(df$Regulator, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Regulator", "Count"),
    caption = "Presence of Independent Regulators"
  ) %>%
  kable_styling(full_width = FALSE)
```

## Private IPP Established
- **Type:** Logical
- **Description:** TRUE if Independent Power Producer (IPP) exist.
- **Source:** World Bank Market Structure Dataset
```{r}
table(df$`Private IPP Established`, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("IPP Established", "Count"),
    caption = "Existence of Independent Power Producers (IPPs)"
  ) %>%
  kable_styling(full_width = FALSE)
```

\newpage
## Unbundling
- **Type:** Logical
- **Description:** TRUE if transmission is unbundled.
- **Source:** World Bank Market Structure Dataset
```{r}
table(df$Unbundling, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Unbundling", "Count"),
    caption = "Transmission Unbundling Status"
  ) %>%
  kable_styling(full_width = FALSE)
```

## Out of VIU
- **Type:** Logical
- **Description:** TRUE if the utility is no longer vertically integrated.
- **Source:** World Bank Market Structure Dataset
```{r}
table(df$`Out of VIU`, useNA = "ifany") %>%
  as.data.frame() %>%
  kable(
    col.names = c("Out of VIU", "Count"),
    caption = "Utilities Exiting Vertically Integrated Utility (Out of VIU)"
  ) %>%
  kable_styling(full_width = FALSE)
```

## Access 2022
- **Type:** Numeric
- **Description:** Percent of national population with electricity access in 2022.
- **Source:** World Bank World Development Indicators (WDI)
```{r}
df %>%
  summarize(
    Min = min(`Access 2022`, na.rm = TRUE),
    Q1 = quantile(`Access 2022`, 0.25, na.rm = TRUE),
    Median = median(`Access 2022`, na.rm = TRUE),
    Mean = mean(`Access 2022`, na.rm = TRUE),
    Q3 = quantile(`Access 2022`, 0.75, na.rm = TRUE),
    Max = max(`Access 2022`, na.rm = TRUE),
    NAs = sum(is.na(`Access 2022`))
  ) %>%
  kable(caption = "Summary - Access to Electricity (2022)", align = "c") %>%
  kable_styling(full_width = FALSE)
```

# Notes

- DEA efficiency scores were calculated using the CCR input-oriented model.
- Monetary figures are in millions of USD.
- Logical fields show TRUE/FALSE flags for market conditions.
- This dataset was developed as part of my Substantial Research Paper (SRP) to fulfill the requirements for the Master of Arts in International Economic Relations: Quantiative Methods at American University. The data preparation, analysis, and documentation were completed in Spring 2025.
